{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2024-12-21 paper.tex
Work on design considerations section

\section{Background}
What is this about? end-user programming \cite{myers-2006-eup}
but also programming systems \cite{jakubovic-2023-techdims},
Live \cite{rein-2019-live}, Interactive etc. - perhaps for a lack of better term use end-user
(also notebooks)

@@

@@

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
\vspace{-1.25em}
\caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
the ``Remove completed'' button (2) modifies target and specifies a condition. }
\label{fig:interactive}
\vspace{-2em}
\end{figure}

@@

The Denicek substrate can be used to generalize the interactions recorded through programming by
demonstration. Our prototype illustrates this option in a limited way. As shown in
Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
In addition to the recorded interaction, we manually specify (in the source view) that the edits
should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
applied to elements for which the formula (which tests if the checkbox is checked) specified by a
relative selector {\small\Verb_./condition/comp/result_} evaluates to true.

@@

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
\vspace{-1em}
\caption{Correcting an incorrect formula. The user uses Copy to create the second
formula (1). They notice an error (2), go back in history to switch left and right argument (3),
merge the change and re-evaluate both formulas (4).}
\label{fig:copypaste}
\vspace{-0.5em}
\end{figure}

@@

\section{Design Discussion}
\label{sec:discuss}

\subsection{Design principles}
\label{sec:discuss-principles}

PRINCIPLES
* merging of edits should not require looking at current document (non-computationality)

\subsection{Design considerations}
\label{sec:discuss-considerations}

\subsubsection*{explicit indices}
DIFFICULTIRES
* if you are not able to refer to newly added items, you cannot reasonably do PBD
  => need non-numerical indices

unifying lists and records?

\subsubsection*{mrdts and ordering}
LISTS vs RECORDS
- are they really different? also MRDT idea - needed for both

% https://gowthamk.github.io/docs/mrdt.pdf page 11

USE MRDTs for list order!

@@

\subsubsection*{Better apply to added}

not so easy because of the non-computationality principle
(we cannot know what will need to happen)

how to transform appends?
* apply to added? (does not work for copy)
* copy to doc and edit? (good because edits are isolated from other things, but messy code!)
(if we edit in doc, it works and we also do not need indices for ListAppend)
* keep nested edits for append op? also does not work, because of the tricky case

Tricky case is:
- add new list item
- then set its fields
  (this breaks things because if we push "add" through structural edits, they will not affect the new values added in the second step)

TODO: Delete ListAppend

AppendFrom - hard to avoid! because recorded edits cannot know length
(but having -1 in index would break too)

\subsubsection*{Schema vs data}
make schema explicit?

@@

\subsection{Heuristic evaluation}
\subsection{Limitations}

@@

VALUE vs STRUCTURE distinction
* good in theory, nice for implementation
* tricky to use! needs some assistance tools

SUBSTRATE DESIGN PROBLEMS
* selectors - all for structure / index for data
  (but it is useful to allow others...)
  (multiselect also bad for checks!)
* groups/conditions/preconditions
  (c.f. email to jonathan)
  tried conditions on edits; trying groups with check edits
* what to do with "disabled edits"? for example when we remove all checked
  (before, this created edit groups with "check" but if the check was false,
  the group was ignored and this messed up merging - because we wouldn't know if the
  edit had any effect or not)

Evaluation
* evaluated edits have to be migrated to the end
  (if there are conflicts, they are dropped)
  Think of this as maintaining a tree:

  e3
  |
  e2   evaluated
  |  /
  e1
  |
  e0

  this has to be serialized as e0 -> e1 -> e2 -> e3 -> evaluated

  evaluated edits do not became part of the main history
  but hang on the side

NOTES
* ListAppendFrom - we need this, because we cannot encode this.
* for records, we can RecordAdd(sel, fld, ..) @ Copy(sel @ [Field fld], src) but
  this does not work for lists - because we do not know the index!
  (and we cannot look into current document, because it will differ for saved-interactions)

TODO
* many things with <tag> selectors currently do not work
  (e.g. `matches` for highlighting) because if we collect path of a current node,
  we collect indices and get /some/2/another - and cannot tell if this matches
  /some/<li>/another - we'd have to collect more detailed path info!

>>>>

\section{Background}
\label{sec:background}

What is this about? end-user programming \cite{myers-2006-eup}
but also programming systems \cite{jakubovic-2023-techdims},
Live \cite{rein-2019-live}, Interactive etc. - perhaps for a lack of better term use end-user
(also notebooks)

@@

\cite{gobert-2023-lorgnette}

The Denicek substrate owes much to systems discussed in \S\ref{sec:background},
especially Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}.

@@

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
\vspace{-1.25em}
\caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
the ``Remove completed'' button (2) modifies target and specifies a condition. }
\label{fig:interactive}
\vspace{-1em}
\end{figure}

@@

Denicek can be used to generalize the interactions recorded through programming by
demonstration. Our prototype illustrates this option in a limited way. As shown in
Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
In addition to the recorded interaction, we manually specify (in the source view) that the edits
should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
applied to elements for which the formula (which tests if the checkbox is checked) specified by a
relative selector {\small\Verb_./condition/comp/result_} evaluates to true.

@@

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
\vspace{-1.5em}
\caption{Correcting an incorrect formula. The user uses Copy to create the second
formula (1). They notice an error (2), go back in history to switch left and right argument (3),
merge the change and re-evaluate both formulas (4).}
\label{fig:copypaste}
\vspace{-0.5em}
\end{figure}

@@

\section{Design Considerations}
\label{sec:discuss}

The design of the Denicek substrate is the result of an iterative process in which we repeatedly
adapted the Denicek design until we obtained a satisfactory solution for the six formative examples
(\S\ref{app:examples}) in the Webnicek prototype. In this section, we document the design challenges,
many of which have until now been personal knowledge of researchers working on related
systems \cite{jakubovic-2022-ladder,edwards-2005-subtext,hall-2017-infra,omar-2021-livelits}.

Two guiding principles for Denicek have been \emph{composability} and
\emph{uniformity}~\cite{jakubovic-2023-techdims}. The design should cover a large number of
cases using a small number of concepts. This is apparent in the design of the document
structure (a node can be data, code or rich text) as well as in the role of edits
(an edit can be a value or structure change, the result of user interaction or the
result of evaluation).

The inconvenience associated with composability and uniformity~\cite{jakubovic-2023-techdims} is not
a problem for Denicek. We view Denicek as an underlying structure on top of which more
convenient end-user programming systems can be built. The uniform design offers a degree of open-endedness,
making it possible to use the basic structures of Denicek in new ways, not anticipated in this paper.

\subsubsection*{Implicit vs. Explicit Structure}
A key design choice for a substrate like Denicek is whether to track the document structure
explicitly or make it implicit. In Denicek, we assume that elements of a list node have the same
structure (i.e.~records should have the same fields, so that they can be targetted using the \ident{All}
selector), but we do not enforce this. Often, the property is temporarily violated when adding a
list item, but then restored once the item is created.

Keeping the structure (type) explicit has theoretical appeal and it simplifies some
aspects of the implementation. In particular, edits to document structure can be clearly separated
from edits of document values. The same separation has disadvantages in that it complicates the
edit language and it requires novel user interface. (In Webnicek, we can select all list items
and edit them at once using the same mechanism as when editing individual list items. With explicit
structure, changing a list structure has to be either a different kind of edit or possibly an edit
of a virtual ``prototype'' element.)

\subsubsection*{Merging and Document State}
One specific design choice in Denicek is that the merge operation operates solely on two
sequences of edits, but it does not need access to the current document state (or full edit
histories). The reconcilliation of edits (\S\ref{sec:system-ops}) is defined for two individual
edits $e_1$ and $e_2$.

This limits the language of edits in a number of ways. The first consequence is that we cannot, in
general, support conditional edits (edits that are applied only when a certain condition holds for
the current document state). This is the case because the merging operation would not be able to
determine whether the edit is applied (and whether its effects should take place). One consequence
of this choice is that generalization of recorded interactions (\S\ref{sec:impl-interaction}) has to
be done on a meta-level (by specifying how to invoke edits, rather than in the edits themselves).
The second consequence of this design choice is that our langauge of selectors has to be abstract.
We can target all list elements using the \ident{All} selector, but cannot convert between this
representation and a representation that explicitly lists indices of all items currently in the list.

\subsubsection*{Explicit List Indices}
Denicek does not use implicit numerical indices for lists. When adding an item, the
user (or the system built on top of Denicek) has to explicitly provide an index. This design
is motivated by a typical programming by demonstration scenario where the user adds a
new list item and then modifies it (adding a speaker or a Todo item). When using
explicit indices, the edits following \ident{Append} can easily refer to the newly added
item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
indices of newly added items in recorded interactions in \S\ref{sec:impl-pbd}).

Using implicit numerical indices is possible, but it requires treating all list-related edits
as operations that transform document selectors. This means adding multiple new rules to
Figure~\ref{fig:updates}, which specifies how edits transform references. Merging then has to
modify indices when edits add, remove or reorder list items. Somewhat, counterintuitively, list
order then becomes a part of document structure rather than a property of the list value.

The problem could also be avoided by not letting the user modify a list item after adding it.
(We can require new items to be constructed outside of the list and then added through a
single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easier to
implement, but it requires unintuitive user interaction and it also does not address all
use cases for modifying list item via index.

\subsubsection*{Ordering List Items and Record Fields}
When list indices are not numerical and supplied explicitly, we need another mechanism for
tracking order of list items. Denicek uses a data structure inspired by Mergeable Replicated
Data Types \cite{kaki-2019-mrdts} and requires specifying the index of a preceding list item
when appending to a list. (If two edits append to the same list independently with the same
preceding item, the order is non-deterministic.)

Note that Denicek uses the same mechanism for record fields. Keeping record fields ordered is
desirable because Denicek documents map directly to HTML documents and so the order is visible
to a user. Moreover, different ways of merging can result in different order of adding fields
to a record. Making order explicit ensures those result in equivalent documents.

Denicek still makes a distinction between records and lists, even if the two are technically
extremely similar. The difference is conceptual. A list is expected to contain items of the
same structure (that can be addressed using the \ident{All} selector), whereas record fields
are expected to be of different types. If we unified lists and records, \ident{All} would have
to be applicable to records too and inferring and checking document structure would be
challenging.

@@

\subsubsection*{Fancy selectors}

* many things with <tag> selectors currently do not work
  (e.g. `matches` for highlighting) because if we collect path of a current node,
  we collect indices and get /some/2/another - and cannot tell if this matches
  /some/<li>/another - we'd have to collect more detailed path info!

  * selectors - all for structure / index for data
    (but it is useful to allow others...)
    (multiselect also bad for checks!)

\subsubsection*{Dependency tracking}

% ---------------------------------

@@

\subsection{Heuristic evaluation}

techdims

\subsection{Limitations}

@@
